---
title: "537 Project"
output: pdf_document
date: "2025-12-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Exploration
```{r}
setwd("~/Desktop/BIS537/project")
data <- read.csv("./block_for_stats.csv")
data$pt_post48_IMV <- ifelse(data$pt_post48_IMV == "True", 1, 0)
data$is_dead_hosp <- ifelse(data$is_dead_hosp == "True", 1, 0)
# all pt_pre24_IMV are 0. 

keep_cols <- c("encounter_block", "age", "sex_category", "mimic_race", "weight_kg", 
               "insurance", "language", "elixhauser_age_adj", "ICU_type", "pt_post48_IMV", 
               "icu_los_days", "is_dead_hosp", "vent_free_days")
df <- data[, keep_cols]
colSums(is.na(df))
df <- df[complete.cases(df[, keep_cols]), ]
```

# Primary Analysis: Doubly-Robust Estimation (AIPW)
## PS Model Estimation Using Standard Logistic Regression
```{r}
ps_model <- glm(pt_post48_IMV ~ age + sex_category + mimic_race + weight_kg + insurance + 
                  language + elixhauser_age_adj + ICU_type, data = df, 
                family = binomial(link = "logit"))
df$pscore <- predict(ps_model, type = "response")
summary(df$pscore)
```


```{r, fig.height=4, fig.width=6}
library(ggplot2)

ggplot(df, aes(x = pscore, fill = as.factor(pt_post48_IMV))) +
  geom_density(alpha = 0.4) +
  labs(title = "Propensity Score Overlap", x = "Estimated Propensity Score",
       fill = "Early PT") +
  theme_minimal()
```

```{r}
library(cobalt)
library(WeightIt)

w <- weightit(pt_post48_IMV ~ age + sex_category + mimic_race + weight_kg +
                insurance + language + elixhauser_age_adj + ICU_type,
              data = df, method = "ps", estimand = "ATE")

bal <- bal.tab(w, s.d.denom = "pooled", un = TRUE, abs = TRUE)

love.plot(bal, stats = "m", abs = TRUE, thresholds = c(m = .10),
          var.order = "unadjusted", line = FALSE,
          colors = c("grey50","red3"), stars = "std",
          title = "Standardized Mean Differences")
```

## PS Model Estimation Using PSweight
```{r}
library(PSweight)

ps_formula <- pt_post48_IMV ~ age + sex_category + mimic_race + weight_kg +
  insurance + language + elixhauser_age_adj + ICU_type

bal_obj <- SumStat(ps.formula = ps_formula, data = df, weight = "IPW")
plot(bal_obj, type = "hist")
```

```{r}
## ASD 
asd_tab <- summary(bal_obj, weighted.var = TRUE, metric = "ASD")
asd_tab
## PSD
psd_tab <- summary(bal_obj, weighted.var = TRUE, metric = "PSD") 
psd_tab
```


```{r, fig.height=6, fig.width=12}
## Love plot with ASD
plot(bal_obj, type = "balance", metric = "ASD", weighted.var = TRUE, threshold = 0.10)
## Love plot with PSD
plot(bal_obj,type = "balance", metric = "PSD", weighted.var = TRUE, threshold = 0.10)
```

```{r, fig.height=4, fig.width=6}
names(bal_obj$ps.weights)
str(bal_obj$ps.weights)

w_ipw <- bal_obj$ps.weights[,"IPW"]  # or bal_obj$ps.weights$IPW
summary(w_ipw)
quantile(w_ipw, probs = c(0.5, 0.9, 0.95, 0.99, 1))
hist(w_ipw, breaks = 50, main = "IPW weight distribution", xlab = "IPW weight")
```

The histogram shows most weights are small and there is no huge right tail.There is no single bar stretching way out to the right, which would indicate a few observations with extremely large weights driving the analysis. From this plot, there’s no obvious weight-instability problem. 


```{r}
A <- df$pt_post48_IMV  # 0/1
sum(df$pt_post48_IMV)

## Overall ESS
ESS_all <- (sum(w_ipw))^2 / sum(w_ipw^2)

## ESS among treated
ESS_treated <- (sum(w_ipw[A == 1]))^2 / sum(w_ipw[A == 1]^2)

## ESS among controls
ESS_control <- (sum(w_ipw[A == 0]))^2 / sum(w_ipw[A == 0]^2)

ESS_all
ESS_treated
ESS_control
```
Original treatment sample is 2753, the sample size of the control group is 27357. 
ESS_trt = 1822.791, ESS_ctr = 27273.81. 
The effective sample sizes remain substantial in both groups. 

**The overlap and covariate balance check are already good enough for PS model using GLM. There is no need to fit a fancier model for PS estimation. **

## ICU LOS Outcome Model Estimation
```{r}
dr_icu_los <- PSweight(
  ps.formula = as.formula("pt_post48_IMV ~ age + sex_category + mimic_race + weight_kg +
                           insurance + language + elixhauser_age_adj + ICU_type"), 
  yname = "icu_los_days", data = df, weight = "IPW",  augmentation = TRUE,  
  out.formula = as.formula("icu_los_days ~ age + sex_category + mimic_race + weight_kg +
                            insurance + language + elixhauser_age_adj + ICU_type"),
  family = "gaussian")

summary(dr_icu_los)
```

## In hospital Mortality Outcome Model Estimation
```{r}
dr_dead_hosp <- PSweight(
  ps.formula = as.formula("pt_post48_IMV ~ age + sex_category + mimic_race + weight_kg +
                           insurance + language + elixhauser_age_adj + ICU_type"), 
  yname = "is_dead_hosp", data = df, weight = "IPW",  augmentation = TRUE,  
  out.formula = as.formula("is_dead_hosp ~ age + sex_category + mimic_race + weight_kg +
                            insurance + language + elixhauser_age_adj + ICU_type"),
  family = "binomial")

summary(dr_dead_hosp)
```

## Ventilator Free Days Outcome Model Estimation
```{r}
dr_vent_free <- PSweight(
  ps.formula = as.formula("pt_post48_IMV ~ age + sex_category + mimic_race + weight_kg +
                           insurance + language + elixhauser_age_adj + ICU_type"), 
  yname = "vent_free_days", data = df, weight = "IPW",  augmentation = TRUE,  
  out.formula = as.formula("vent_free_days ~ age + sex_category + mimic_race + weight_kg +
                            insurance + language + elixhauser_age_adj + ICU_type"),
  family = "gaussian")

summary(dr_vent_free)
```

# Sensitivity Analysis I: Outcome Model using Machine Learning
```{r}
df$sex_category <- factor(df$sex_category)
df$mimic_race <- factor(df$mimic_race)
df$insurance <- factor(df$insurance)
df$language <- factor(df$language)
df$ICU_type <- factor(df$ICU_type)

ps_formula  <- pt_post48_IMV ~ age + sex_category + mimic_race + weight_kg +
                               insurance + language + elixhauser_age_adj + ICU_type
```


## ICU LOS 

We use the default values for other model parameters such as the num- ber of trees (n.trees = 100), interaction depth (interaction.depth = 1), the minimum number of observations in the terminal nodes (n.minobsinnode = 1), shrinkage reduction
29 (shrinkage = 0.1), and bagging fraction (shrinkage = 0.5). 

```{r}
out_formula <- icu_los_days ~ age + sex_category + mimic_race + weight_kg +
                               insurance + language + elixhauser_age_adj + ICU_type

dr_icu_los_gbm <- PSweight(ps.formula = ps_formula, yname = "icu_los_days", data = df,
                              weight = "IPW", augmentation = TRUE, out.formula  = out_formula,
                              family = "gaussian", ps.method = "glm", out.method = "gbm",
                              out.control = list(distribution= "gaussian", n.trees = 1000,
                                                interaction.depth = 2, shrinkage = 0.01,
                                                bag.fraction = 0.5))
summary(dr_icu_los_gbm)
```

## In Hospital Mortality
```{r}
out_formula <- is_dead_hosp ~ age + sex_category + mimic_race + weight_kg +
                               insurance + language + elixhauser_age_adj + ICU_type

dr_is_dead_hosp_gbm <- PSweight(ps.formula = ps_formula, yname = "is_dead_hosp", data = df,
                              weight = "IPW", augmentation = TRUE, out.formula  = out_formula,
                              family = "binomial", ps.method = "glm", out.method = "gbm",
                              out.control = list(distribution= "bernoulli", n.trees = 1000,
                                                interaction.depth = 2, shrinkage = 0.01,
                                                bag.fraction = 0.5))
summary(dr_is_dead_hosp_gbm)
```

## Ventilator Free Days
```{r}
out_formula <- vent_free_days ~ age + sex_category + mimic_race + weight_kg +
                               insurance + language + elixhauser_age_adj + ICU_type

dr_vent_free_gbm <- PSweight(ps.formula = ps_formula, yname = "vent_free_days", data = df,
                              weight = "IPW", augmentation = TRUE, out.formula  = out_formula,
                              family = "gaussian", ps.method = "glm", out.method = "gbm",
                              out.control = list(distribution= "gaussian", n.trees = 1000,
                                                interaction.depth = 2, shrinkage = 0.01,
                                                bag.fraction = 0.5))
summary(dr_vent_free_gbm)
```

# Evaluation of Point Estimation: Visualization 
```{r}
library(dplyr)
library(ggplot2)

res <- tribble(
  ~outcome,            ~method,        ~estimate,   ~lci,       ~uci,
  # GLM outcome models
  "Vent-free days",    "GLM outcome",   1.34178,     0.86315,   1.82040,
  "In-hosp mortality", "GLM outcome",  -0.0535613,  -0.0702377,-0.036885,
  "ICU LOS",           "GLM outcome",  -0.13846,    -0.48743,   0.21051,
  # GBM outcome models
  "Vent-free days",    "GBM outcome",   1.29565,     0.82559,   1.76570,
  "In-hosp mortality", "GBM outcome",  -0.0512948,  -0.0682490,-0.034341,
  "ICU LOS",           "GBM outcome",  -0.097394,   -0.436632,  0.24184
)

res$outcome <- factor(res$outcome, levels = c("ICU LOS", "In-hosp mortality", "Vent-free days"))

ggplot(res, aes(x = outcome, y = estimate, color = method, group = method)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(position = position_dodge(width = 0.4), size = 3) +
  geom_errorbar(aes(ymin = lci, ymax = uci), width = 0.1, 
                position = position_dodge(width = 0.4)) +
  labs(x = "Outcome", y = "Estimated treatment effect (Contrast 1)", color = "Outcome model") +
  theme_bw() +
  theme(axis.title.x = element_text(margin = margin(t = 8)))
```


